Evaluating the Performance and Energy Efficiency of COSMO-ART, a Coupled Numerical Weather Forecast and Chemical Transport Model.

### 2014-04-22 paper_enahpc14/Introduction.tex
commits Will's slight corrections

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  these  uncertainties,  the  research  community  is  making  a
concerted international  effort to represent  the underlying physical,
chemical  and aerosol dynamical  processes through  numerical chemical
transport  models (CTMs)  such  as ART  (Aerosols  and Reactive  Trace
gases),   developed   at  the   Karlsruhe   Institute  of   Technology
(KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
scale  model system  is online  coupled with  the  operational weather
forecast  model  COSMO \citep{Baldauf-2011},  jointly  developed by  a
consortium of European weather centers and used in the climate version
(COSMO-CLM) by a wide research community.

The extended  COSMO-ART model  is computationally much  more demanding
than the  COSMO core  since a large  number of additional  tracers and
processes  have  to  be  considered.   Thus this  model  is  currently
severely-limited in  terms of applicability and expensive  in terms of
energy consumption.   Although COSMO has recently been  ported to GPUs
within  the framework of  the High  Performance and  High Productivity
Computing  (HP2C)  Initiative to  optimize  it  for computational  and
energy efficiency (\url{http://www.hp2c.ch/}), significant investments
in  ART  are still  required  to  take it  to  a  similar level.   The
efficiency of COSMO-ART is being addressed in the EU Exa2Green project
(\url{http://exa2green.eu/}). The  ultimate goal of the  project is to
deliver a  prototype code, which  provides an energy efficiency  of at
least five times of the  baseline value.  Such an implementation would
allow  the  community  to  investigate critical  questions  at  higher
resolution  and   over  longer  periods,   at  reduced  cost   to  the
environment.

Some  related work  will  be discussed  in  Sec.~\ref{sec:1}. Then  in
Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
COSMO-ART and specify its  technical setup related to the investigated
performance and energy evaluation methods.  Sec.~\ref{sec:3} describes
different  power  measurement  equipments  used  for  conducting  this
benchmarking  study and presents  energy-aware techniques  employed to
determine  the   energy  footprint  of   the  considered  application.
Sec.~\ref{sec:4} gathers benchmark  results from the Exa2Green project
and  highlights areas  where improvements  will be  necessary  for the
subsequent baseline refactoring.   Finally, we conclude with practical
guidelines  to identifying  energy  saving opportunities  and give  an
outlook to future research in Sec.~\ref{concl}.

>>>>

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  these  uncertainties,  the  research  community  is  making  a
concerted international  effort to represent  the underlying physical,
chemical  and aerosol dynamical  processes through  numerical chemical
transport  models (CTMs)  such  as ART  (Aerosols  and Reactive  Trace
gases),   developed   at  the   Karlsruhe   Institute  of   Technology
(KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
scale model  system is coupled  with the operational  weather forecast
model COSMO \citep{Baldauf-2011}, jointly developed by a consortium of
European weather  centers and used in the  climate version (COSMO-CLM)
by a wide research community.

The extended  COSMO-ART model  is computationally much  more demanding
than COSMO  since a large  number of additional tracers  and processes
have to  be considered.  Thus the model  is currently severely-limited
in  terms   of  applicability  and   expensive  in  terms   of  energy
consumption.    Although   COSMO   has   recently   been   ported   to
GPUs  \citep{Gysi-2014, Lapillonne-2014} within  the framework  of the
High Performance and High  Productivity Computing (HP2C) Initiative to
optimise    it    for     computational    and    energy    efficiency
(\url{http://www.hp2c.ch/}), significant investments  in ART are still
required to take  it to a similar level.   The efficiency of COSMO-ART
is     being    addressed    in     the    EU     Exa2Green    project
(\url{http://exa2green.eu/}). The  ultimate goal of the  project is to
deliver a  prototype code, which  provides an energy efficiency  of at
least five times of the  baseline value.  Such an implementation would
allow  the  community  to  investigate critical  questions  at  higher
resolution  and   over  longer  periods,   at  reduced  cost   to  the
environment.

This  work is organized  as follows:  in Sec.~\ref{sec:1}  we describe
related  work,  then in  Sec.~\ref{sec:2},  we  briefly introduce  the
regional atmospheric  model COSMO-ART and specify  its technical setup
related to the investigated performance and energy evaluation methods.
Sec.~\ref{sec:5} describes the HPC  supercomputers used to compute the
baseline.  Sec.~\ref{sec:3}   describes  different  power  measurement
equipment  used for  conducting this  benchmarking study  and presents
energy-aware techniques employed to  determine the energy footprint of
the   considered  application.   Sec.~\ref{sec:4}   gathers  benchmark
results  from  the  Exa2Green   project  and  highlights  areas  where
improvements   will   be  necessary   for   the  subsequent   baseline
refactoring.   Finally,  we  conclude  with  practical  guidelines  to
identifying energy saving opportunities  and give an outlook to future
research in Sec.~\ref{concl}.

### 2014-04-22 paper_enahpc14/Sec2:COSMO-ART.tex
commits Will's slight corrections

\subsection{Model Setup}
\label{subsec:1.2}
Establishing effective energy performance benchmarking of a code under
intense development such as COSMO-ART is a challenging task because of
the  absolute necessity that  results must  be reproducible  within an
expected variance for the duration of the Exa2Green project. To define
a baseline,  it was necessary  to find a run-configuration  capable of
being  recreated in  lal subsequent  versions.  Here  we give  a brief
overview of the model setup  for our performance and energy efficiency
evaluation.

@@

The  considered  COSMO-ART model  system  is  configured  with a  semi
Lagrangian horizontal advection scheme with tricubic interpolation and
selective filling  diffusion option in combination  with the dynamical
core  using Runge-Kutta  time  stepping \citep{COSMO-PartI-2011}.   It
also  makes use  of  the  Kinetic PreProcessor  solver  (KPP) for  the
resolution  of atmospheric  chemistry ordinary  differential equations
\citep{Damian-2002}.   Concerning the modelling  of wet  deposition in
aerosols, the  baseline has only indirect cloud  feedbacks but does not
include  in-cloud  scavenging  (rainout)  and  below-cloud  scavenging
(washout)  yet.   Amongst  physical  parameterisations,  precipitation
formation is performed by a two-moment cloud microphysics.

>>>>

\subsection{Model Setup}
\label{subsec:1.2}
Establishing effective energy performance benchmarking of a code under
intense development such as COSMO-ART is a challenging task because of
the  absolute necessity that  results must  be reproducible  within an
expected variance for the duration of the Exa2Green project. To define
a baseline,  it was necessary  to find a run-configuration  capable of
being  recreated in  all subsequent  versions.  Here  we give  a brief
overview of the model setup  for our performance and energy efficiency
evaluation.

@@

The   considered  COSMO-ART   model  system   is  configured   with  a
semi-Lagrangian    horizontal   advection    scheme    with   tricubic
interpolation  and selective filling  diffusion option  in combination
with   the   dynamical    core   using   Runge-Kutta   time   stepping
\citep{COSMO-PartI-2011}.    It  also   makes  use   of   the  Kinetic
PreProcessor solver (KPP) for  the resolution of atmospheric chemistry
ordinary  differential equations \citep{Damian-2002}.   Concerning the
modelling  of  wet  deposition  in  aerosols, the  baseline  has  only
indirect  cloud feedbacks  but  does not  include in-cloud  scavenging
(rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
parameterisations,   precipitation  formation   is   performed  by   a
two-moment cloud microphysics.

### 2014-04-22 paper_enahpc14/Sec4:Evaluation.tex
cancels previous commits of Cristiano as IBM Zurich is actually not going to contribute to this work
commits Will's slight corrections

\subsection{Environment setup}
\label{subsec:4.2}
While hardware  platforms mature  and are replaced,  we have  chosen a
state-of-the-art  Intel's  third-generation   Core  (aka  Ivy  Bridge)
processing platform (called ``Monch'')  for our power measurements, as
it  is slated  to stay  in service  without hardware  upgrade  for the
duration  of  the Exa2Green  project.   In  principle  at least,  this
architecture could be recreated or found in an identical configuration
beyond the lifetime of the project.  Given that the baseline benchmark
can be reproduced  within an expected variance, and  that the baseline
run configuration  can be used in  all future versions of  the code, a
fair comparison  will be made  between the baseline and  the milestone
versions   of    COSMO-ART.    In   this    study,   a   complementary
energy-to-solution  benchmarking  comparison  is  carried out  on  the
``Pilatus'' cluster based  on Intel's previous-generation Sandy Bridge
processors, conventional  in HPC  systems and known  to be  more power
consuming.   Finally a  full tracing  experiment is  conducted  on the
``Tintorrum''  cluster using  an integrated  framework, to  capture an
overall  power profile  at a  much finer  resolution and  to  get more
insights  into  the  MPI  blocking  and polling  influences  on  power
savings.

\subsubsection{Monch (CSCS - ETH Zurich)}
Monch, which is installed  at the Swiss National Supercomputing Center
(CSCS) of ETH Zurich, is  a 10 rack NEC-provided and dual-socket Intel
Ivy Bridge-EP based cluster, utilised from people that are part of the
Swiss    Platform   for    Advanced   Scientific    Computing   (PASC,
\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
Each standard  compute node comprises  two Intel Ivy  Bridge Efficient
Performance (EP)  E5-2660 v2 ten-core processors operating  at 2.2 GHz
base  clock speed,  themselves connected  by a  high  speed InfiniBand
network based on Mellanox SX6036  managed FDR switches, with a 56 Gb/s
speed.  Each core  has 32 KB instruction and 32 KB  data L1 caches and
256 KB  of L2 cache. All  the 10 core share  a 25 MB L3  cache and the
platform has  32GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
benchmark, a  full rack  of Monch constituted  of 52  standard compute
nodes (monchc[029-080]) was considered.

@@

\subsubsection{Sofware environment}
INT2LM  and the  COSMO-ART model  are  implemented in  Fortran 90  for
distributed  memory  parallel  computers  using  the  Message  Passing
Interface (MPI)  and are purely  MPI-parallel.  The software  stack on
both  many-core  CSCS  platforms  was  controlled  using  the  modules
framework which gives an easy  and flexible mechanism to access to all
of the  provided compilers, tools  and applications.  For  our initial
benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,
gcc/4.8.2 on Pilatus) with the -O3 compiler flag as it generally gives
a  good  level  of optimization  and  the  code  runs faster  in  this
configuration  than when  compiled with  the intel  compiler (14.0.1).
Besides, we installed the  MPICH2 implementation of MPI (mvapich2/1.9)
as  well  as  the  commonly  used HDF5  (1.8.12)  and  NetCDF  (4.3.1)
libraries in favor of the  traditional GRIB library for the management
of extremely  large and complex data collections.   All computes nodes
have    an   operating   system    based   on    GNU/Linux   featuring
``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
``3.0.101-0.15-default'' kernel in Pilatus.

\subsubsection{Run configuration}
A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
Pilatus.   The  calculated  region  was mapped  to  the  participating
processors using  a 2D-partitioning strategy.   The distribution along
the  x and  y  coordinates  was defined  by  setting: $nprocx=40$  and
$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
usually  kept  bigger  than  $nprocy$.   Besides as  this  version  of
COSMO-ART  doesn't  make  use   of  the  GRIB  library,  we  specified
$nprocio=0$ for  GRIB I/O.  Hyperthreading  is not considered  in this
study as previous  attempts of its use revealed that  it always led to
higher energy-to-solution.\\

@@

In Figure~\ref{fig:3}, we compare both time-to-solution (right y-axis)
and  energy-to-solution (left  y-axis) metrics  on both  platforms. As
expected,  Xeon  E5 outperforms  Ivy  Bridge  EP,  being roughly  1.3x
faster.  The reason for that is twofold: it has higher clock frequency
than Xeon E5 (2.6 GHz against  2.2 GHz) and it aims at computing speed
regardless to  energy consumption. In  our experiments, Ivy  Bridge EP
showed the best energy-to-solution, reducing the energy consumption of
Xeon E5 by approximately $7\%$.

>>>>

\subsection{Environment setup}
\label{subsec:4.2}
We have  chosen a state-of-the-art Intel's  third-generation Core (aka
Ivy  Bridge)  processing platform  (called  ``Monch'')  for our  power
measurements,  which is  slated to  stay in  service  without hardware
upgrade for  the duration of  the Exa2Green project.  In  principle at
least, this architecture  could be recreated or found  in an identical
configuration beyond the lifetime  of the project.  Since the baseline
benchmark can be reproduced within  an expected variance, and that the
baseline run configuration  can be used in all  future versions of the
code,  a fair comparison  will be  made between  the baseline  and the
milestone  versions  of COSMO-ART.   In  this  study, a  complementary
energy-to-solution  benchmarking  comparison  is  carried out  on  the
``Pilatus'' cluster based  on Intel's previous-generation Sandy Bridge
processors, conventional  in HPC  systems and known  to be  more power
consuming.   Finally a  full tracing  experiment is  conducted  on the
``Tintorrum''  cluster using  an integrated  framework, to  capture an
overall  power profile  at a  much finer  resolution and  to  get more
insights  into  the  MPI  blocking  and polling  influences  on  power
savings.

\subsubsection{Monch (CSCS - ETH Zurich)}
Monch, which is installed  at the Swiss National Supercomputing Center
(CSCS) of ETH Zurich, is  a 10 rack NEC-provided and dual-socket Intel
Ivy Bridge-EP  based cluster, utilised  by scientists who are  part of
the   Swiss  Platform   for  Advanced   Scientific   Computing  (PASC,
\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
Each standard  compute node comprises  two Intel Ivy  Bridge Efficient
Performance (EP)  E5-2660 v2 ten-core processors operating  at 2.2 GHz
base  clock speed,  themselves connected  by a  high  speed InfiniBand
network based on Mellanox SX6036  managed FDR switches, with a 56 Gb/s
speed.  Each core  has 32 KB instruction and 32 KB  data L1 caches and
256  KB of L2  cache. All  10 cores  share a  25 MB  L3 cache  and the
platform has  32GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
benchmark, a  full rack  of Monch constituted  of 52  standard compute
nodes (monchc[029-080]) was considered.

@@

\subsubsection{Sofware environment}
INT2LM  and the  COSMO-ART model  are  implemented in  Fortran 90  for
distributed  memory  parallel  computers  using  the  Message  Passing
Interface (MPI)  and are purely  MPI-parallel.  The software  stack on
both  many-core  CSCS  platforms  was  controlled  using  the  modules
framework which gives an easy  and flexible mechanism to access to all
of the  provided compilers, tools  and applications.  For  our initial
benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,
gcc/4.8.2 on Pilatus) with the -O3 compiler flag as it generally gives
a  good  level  of optimisation  and  the  code  runs faster  in  this
configuration  than when  compiled with  the intel  compiler (14.0.1).
Besides, we installed the  MPICH2 implementation of MPI (mvapich2/1.9)
as  well  as  the  commonly  used HDF5  (1.8.12)  and  NetCDF  (4.3.1)
libraries in favor of the  traditional GRIB library for the management
of extremely  large and complex data collections.   All computes nodes
have    an   operating   system    based   on    GNU/Linux   featuring
``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
``3.0.101-0.15-default'' kernel in Pilatus.

\subsubsection{Run configuration}
A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
Pilatus.   The  calculated  region  was mapped  to  the  participating
processors using  a 2D-partitioning strategy.   The distribution along
the $x$  and $y$ coordinates  was defined by setting:  $nprocx=40$ and
$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
usually  kept  bigger than  $nprocy$.   Besides,  as  this version  of
COSMO-ART  doesn't  make  use   of  the  GRIB  library,  we  specified
$nprocio=0$ for  GRIB I/O.  Hyperthreading  is not considered  in this
study as previous  attempts of its use revealed that  it always led to
higher energy-to-solution.\\

@@

In Figure~\ref{fig:3}, we compare both time-to-solution (right y-axis)
and  energy-to-solution (left  y-axis) metrics  on both  platforms. As
expected,  Xeon  E5 outperforms  Ivy  Bridge  EP,  being roughly  1.3x
faster.   The  reason  for  that  is two-fold:  it  has  higher  clock
frequency than  Ivy Bridge (2.6  GHz against 2.2  GHz) and it  aims at
computing speed regardless to  energy consumption. In our experiments,
Ivy Bridge EP showed  the best energy-to-solution, reducing the energy
consumption of Xeon E5 by approximately $7\%$.
